---
title: "Introduction to microFGT"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to microFGT}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

The microFGT package provides specialized tools for analyzing microbiome data from the female genital tract (FGT). It is built on the TreeSummarizedExperiment data structure from Bioconductor and includes functions for data manipulation, visualization, and specialized FGT-specific analyses.

```{r setup}
library(microFGT)
```

## Basic Usage

### Creating FGTExperiment Objects

The core data structure in microFGT is the FGTExperiment class, which extends TreeSummarizedExperiment. You can create an FGTExperiment object from raw count data:

```{r fgt-experiment}
# Create a simple count matrix
counts <- matrix(sample(0:100, 60, replace = TRUE), nrow = 10, ncol = 6)
rownames(counts) <- paste0("Feature", 1:10)
colnames(counts) <- paste0("Sample", 1:6)

# Create a basic FGTExperiment
fgt_exp <- FGTExperiment(assays = list(counts = counts))

# View the object
fgt_exp
```

### Working with Example Data

microFGT includes example data for testing and demonstration:

```{r example-data, eval=FALSE}
# Load example data
example_data <- load_example_data()

# Generate synthetic data
synthetic_data <- generate_fgt_example_data(
  n_samples = 10,
  n_features = 50,
  community_types = c("CST-I", "CST-III", "CST-IV")
)
```

### Data Transformation

Transform count data to different abundance metrics:

```{r data-transform}
# Transform to relative abundance
fgt_rel <- transform_abundance(fgt_exp, type = "relative")

# Check the first few values of the relative abundance matrix
head(assays(fgt_rel)$relative[, 1:3])
```

### Filtering Taxa

Filter taxa based on prevalence and abundance:

```{r filter-taxa}
# Filter taxa
fgt_filtered <- filter_taxa(fgt_exp, min_prevalence = 0.5, min_abundance = 0.01)

# Compare dimensions before and after filtering
dim(fgt_exp)
dim(fgt_filtered)
```

### Visualization

Create taxonomic composition plots:

```{r taxa-plot, eval=FALSE}
# Create taxonomy data
taxonomy <- data.frame(
  Phylum = sample(c("Firmicutes", "Bacteroidetes", "Proteobacteria"), 10, replace = TRUE),
  Genus = sample(c("Lactobacillus", "Gardnerella", "Prevotella"), 10, replace = TRUE),
  row.names = rownames(counts)
)

# Add taxonomy to the object
rowData(fgt_rel) <- taxonomy

# Plot taxonomic composition
plot_taxa_composition(fgt_rel, rank = "Phylum")
```

## Next Steps

This vignette provides a basic introduction to microFGT. For more detailed information, please see the package documentation and additional vignettes:

- Importing data from common formats
- Working with taxonomic data
- Visualization techniques
- FGT-specific analyses

---

**Note**: microFGT is in early development (alpha stage). APIs may change in future versions.